### 2015-09-04 aphids/aphids.tex
spell checked

To enable VLBI observations with heteregeneous arrays, it is sometimes necessary to pre-process
the data before correlating.  In this chapter we describe a software solution to change the 
samping rate of the data taken at the SMA for the EHT's March 2015 observing campaign.  We 
focus on the sampling rate conversion problem and investigate several possible algorithms.  The project
has applications for future VLBI observations at the SMA as well as other sites with similar 
challenges.

@@

Sample rate conversion, or resampling, is the process of taking a digital signal ($x[i]$) sampled at some rate 
($f_0$) and calculating new samples ($y[i]$) at a different rate ($f_1$).  To analyze this problem we use the 
ratio between the two rates: $f_1/f_0 = L/M$, where $L$ and $M$ are relatively prime integers.  A sample rate 
conversion can be helpfully thought of as the combination of upsampling and downsampling 
\citep{oppenheim10,lyons11}.  
Upsampling, (also commonly called interpolation or expansion), increases the sampling rate 
by a factor of $L$ with the insertion of $L-1$ zeroes between the original $x[i]$.  A low-pass, 
anti-imaging filter smooths the signal and supresses the high frequency spectral images greater than the original 
$f_0$ that have been introduced by the zero inserts.  Figure \ref{fig:upsampling} sketches this process in the 
time and frequency domain for when $L$ equals 3.

@@

Trimming or padding the spectrum is equivalent to convolving a normalized sinc function with the original series 
wrapped periodically.  Consequenty, errors will be introduced into the beginning and end of the
resampled signal.  How far and significantly this error propagates away from the time series' edges depends 
on the width of the sinc function which is set by the resampling factors $L$ and $M$: 
\begin{equation} \label{eq:sinc}
B\,\mathrm{sinc}(B t) \overset{\mathcal{F}}{\Longleftrightarrow} \begin{cases} 1 \quad |f| < B/2 \\ 0 \quad |f| > B/2 \end{cases}
\end{equation}
where $B = \mathrm{min}(f_0,L/Mf_0)$.  
The number of samples for which this makes a large difference is small and can be roughly treated as less than a 
1\% effect when $N > 100\,\mathrm{max}(L/M,1)$.  One option to mitigate this error to stitch together 
overlapping, resampled segments \citep{bi11} which requires more computation and memory.  A second strategy is to 
multiply the original signal by some tapered window function \citep{fraser89}.

@@

We are developing a software solution to the resampling problem called APHIDS.  This 
open-source project is written in C/CUDA using
the High Availibility Shared Pipeline 
(HASHPIPE)\footnote{\url{https://github.com/david-macmahon/hashpipe}}.  The software is designed to use 
modularized, lightweight threads to handle reading data packets from hard drives, resampling on a GPU cluster, 
and then writing back to separate disks.  These three steps are each loosely mapped to threads we call the 
vdif\_in\_net thread, vdif\_inout\_gpu\_thread, and vdif\_out\_net thread. 
The ultimate goal of is to process data streamed directly at the telescope.  The flowchart in Figure 
\ref{fig:aphids_flow_chart} shows how data progresses 
through the pipeline and this section will describe each block covering the major hardware components, software 
design, and data formats.  We will conclude with a discussion of future development and applications of APHIDS.

@@

APHIDS features three main hardware components: an I/O data system, CPU, and GPU cluster.  For I/O, the 
EHT observations taken by SWARM in March 2015 have been stored on four disk-packs that each hold eight, 
6\,TB hard drives (totaling 192\,TB of disk space).  The Mark6 VLBI data system has four modules 
that can hold one disk-pack and we use
two seprate Mark6 units to handle the recording and playback of data.  APHIDS itself is run on a separate CPU 
that controls the GPU cluster.

@@

APHIDS resamples the data in the Fourier domain (see \S \ref{sec:fourier_resampling}).  There were several 
contributing factors that led us to make
this choice rather than apply filters in the time domain.  First, the GPU cluster can handle relatively large 
chunks of data at a time so losses from the DFT are very manageable.  Second, regardless of 
what resampling algorithm is used the SWARM spectra must 
be transformed into the time domain, the guard bands trimmed and the signal modulated.  These operations are done 
naturally in the Fourier domain.  However, the cost of the resampling operation will be largely dominated by 
these operations.  Lastly, we can utilize the optimized cuFFT library
\footnote{\url{https://developer.nvidia.com/cuFFT}} to ensure good performace, reduce development time and 
maintain a manageably sized codebase.

The APHIDS resampling block is composed of three 1-dimensional FFTs (A, B, and C; see Figure 
\ref{fig:resampling_block}).  Following the cuFFT API reference to improve performace, we utilize batched
FFTs (launching more than one FFT at a time using the cuFFT API), out-of-place transforms, 
and operate on contiguous chunks of memory.  We also use single precision transforms to 
reduce the memory transfer bandwidth and ensure that 
the resampling block can be run on a single device.  The latter point is important since it allows the inout
thread to parallelize the resampling operation across the cluster and provide a roughly x4 speedup.

The first FFT (A) takes the complex valued, 16384 sample SWARM spectrum into the time domain, returning 
32768 real samples.  Since SWARM omits the Nyquist sample which is outside the SWARM guard band cutoff frequency, 
we insert a zero as the 16385th sample.  Since this transform is radix-2, the performance should benefit from 
the Cooley-Tukey algorithm.  However, on our cluster the cuFFT library exhibits linear improvement in Gflops 
with transform size up to N of 16384 (see Figure \ref{fig:C2R_performance}).  This behavior is consistent with 
the CUDA 7.0 performance 
report\footnote{\url{http://on-demand.gputechconf.com/gtc/2015/webinar/gtc-express-cuda7-performance-overview.pdf}} 
and we find that this transform, just ouside of the optimal regime, operates at 6.7 Gsps.

@@

Next, we Fourier transform chunks of $N_B = 19968$ consecutive time series samples (B).  The resulting 
spectrum has a frequency resolution of $2496$\,MHz$/19958=125$\,kHz.  The folllowing step is to apply a third
complex-to-real transform (C) on a subset of the spectrum, using only frequences from $150$\,MHz to $1174$\,MHz.  
The final time series has size $N_C= 2\times(1174-150)/0.125=16384$ per batch element and our desired 
sampling rate of 
$16384 \times 0.125$\,MHz$ = 2048$\,MHz.  This simple operation simultaneously resamples the signal, modulates 
the signal, and trims the guard bands.  Furthermore, it is easily implemented using cuFFT by using the same 
batch size as the second real-to-complex transform, increasing the input pointer index by $150 / 0.125 = 1200$
(equivalent to masking out the first 1200 channels) and then using a input stride of $9980$ for the 
$N=16384$ complex-to-real transform.

@@

\subsection{FFT Accuracy}
One concern with using multiple FFTs to change the sampling rate is the error introduced by bit growth.  The 
floating point error grows as $\mathcal{O}(\sqrt{\log{N}})$ \citep{schatzman96} on average for the Cooley-Tukey
algorithm, but errors can be exacerbated by innacurate twiddle factors or our choice of FFT algorithm (see
benchFFT\footnote{\url{http://www.fftw.org/accuracy/}}).  We have computed the accuracy of the cuFFT library 
using single precision for 
radix-2 transforms as well as the $N_b = 19968$ case in Figure \ref{fig:fft_accuracy}.  These values 
were computed using benchFFT and we compare against fftw3\footnote{\url{http://www.fftw.org/}}
results also computed on our GPU cluster.  For random input, the cuFFT library follows the expected 
$\mathcal{O}(\sqrt{\log{N}})$ behavior, but with larger scatter and slightly worse performance than fftw3.

@@

APHIDS promises to be an important tool for the EHT.  It crucially enables the correlation in
DIFX of non-standard sampling rates and also provides a platform for any general pre-processing of 
VLBI data.  It features fast data I/O from Mark6 recorders and GPU accelarated processing.  
We have demonstrated the correctness and speed of this design and its immediate applicability to 
the 2015 EHT data.  However,
there remain a number of challenges for future development of the software.

For future VLBI campaigns at the SMA, it would be ideal if APHIDS could run in real-time 
at the backend of the SDBE.  This would reduce the media requirements and also the 
hassle of post-processing the data offsite.  However, the performance will need to improve as the
SWARM data rates increase and if the front-end of the pipeline is optimized to reduce the 
amount of SNR loss from quantization.  Additionally, some analysis on the cooling requiremeents 
of the cards should be conducted to ensure that the GTX980s used in our GPU cluster can safely run 
at the altitude of the SMA (4100\,meters).  Regardless of where it operates, APHIDS will need 
maintence to remain compatible with SWARM as it and the SDBE are updated.  The current 
git repository and its wiki is a good solution for keeping track of versioning and performance status.

>>>>

To enable VLBI observations with heterogeneous arrays, it is sometimes necessary to pre-process
the data before correlating.  In this chapter we describe a software solution to change the 
sampling rate of the data taken at the SMA for the EHT's March 2015 observing campaign.  We 
focus on the sampling rate conversion problem and investigate several possible algorithms.  The project
has applications for future VLBI observations at the SMA as well as other sites with similar 
challenges.

@@

Sample rate conversion, or resampling, is the process of taking a digital signal ($x[i]$) sampled at some rate 
($f_0$) and calculating new samples ($y[i]$) at a different rate ($f_1$).  To analyze this problem we use the 
ratio between the two rates: $f_1/f_0 = L/M$, where $L$ and $M$ are relatively prime integers.  A sample rate 
conversion can be helpfully thought of as the combination of upsampling and downsampling 
\citep{oppenheim10,lyons11}.  
Upsampling, (also commonly called interpolation or expansion), increases the sampling rate 
by a factor of $L$ with the insertion of $L-1$ zeroes between the original $x[i]$.  A low-pass, 
anti-imaging filter smooths the signal and suppresses the high frequency spectral images greater than the original 
$f_0$ that have been introduced by the zero inserts.  Figure \ref{fig:upsampling} sketches this process in the 
time and frequency domain for when $L$ equals 3.

@@

Trimming or padding the spectrum is equivalent to convolving a normalized sinc function with the original series 
wrapped periodically.  Consequently, errors will be introduced into the beginning and end of the
resampled signal.  How far and significantly this error propagates away from the time series' edges depends 
on the width of the sinc function which is set by the resampling factors $L$ and $M$: 
\begin{equation} \label{eq:sinc}
B\,\mathrm{sinc}(B t) \overset{\mathcal{F}}{\Longleftrightarrow} \begin{cases} 1 \quad |f| < B/2 \\ 0 \quad |f| > B/2 \end{cases}
\end{equation}
where $B = \mathrm{min}(f_0,L/Mf_0)$.  
The number of samples for which this makes a large difference is small and can be roughly treated as less than a 
1\% effect when $N > 100\,\mathrm{max}(L/M,1)$.  One option to mitigate this error to stitch together 
overlapping, resampled segments \citep{bi11} which requires more computation and memory.  A second strategy is to 
multiply the original signal by some tapered window function \citep{fraser89}.

@@

We are developing a software solution to the resampling problem called APHIDS.  This 
open-source project is written in C/CUDA using
the High Availability Shared Pipeline 
(HASHPIPE)\footnote{\url{https://github.com/david-macmahon/hashpipe}}.  The software is designed to use 
modularized, lightweight threads to handle reading data packets from hard drives, resampling on a GPU cluster, 
and then writing back to separate disks.  These three steps are each loosely mapped to threads we call the 
vdif\_in\_net thread, vdif\_inout\_gpu\_thread, and vdif\_out\_net thread. 
The ultimate goal of is to process data streamed directly at the telescope.  The flowchart in Figure 
\ref{fig:aphids_flow_chart} shows how data progresses 
through the pipeline and this section will describe each block covering the major hardware components, software 
design, and data formats.  We will conclude with a discussion of future development and applications of APHIDS.

@@

APHIDS features three main hardware components: an I/O data system, CPU, and GPU cluster.  For I/O, the 
EHT observations taken by SWARM in March 2015 have been stored on four disk-packs that each hold eight, 
6\,TB hard drives (totaling 192\,TB of disk space).  The Mark6 VLBI data system has four modules 
that can hold one disk-pack and we use
two separate Mark6 units to handle the recording and playback of data.  APHIDS itself is run on a separate CPU 
that controls the GPU cluster.

@@

APHIDS resamples the data in the Fourier domain (see \S \ref{sec:fourier_resampling}).  There were several 
contributing factors that led us to make
this choice rather than apply filters in the time domain.  First, the GPU cluster can handle relatively large 
chunks of data at a time so losses from the DFT are very manageable.  Second, regardless of 
what resampling algorithm is used the SWARM spectra must 
be transformed into the time domain, the guard bands trimmed and the signal modulated.  These operations are done 
naturally in the Fourier domain.  However, the cost of the resampling operation will be largely dominated by 
these operations.  Lastly, we can utilize the optimized cuFFT library
\footnote{\url{https://developer.nvidia.com/cuFFT}} to ensure good performance, reduce development time and 
maintain a manageably sized codebase.

The APHIDS resampling block is composed of three 1-dimensional FFTs (A, B, and C; see Figure 
\ref{fig:resampling_block}).  Following the cuFFT API reference to improve performance, we utilize batched
FFTs (launching more than one FFT at a time using the cuFFT API), out-of-place transforms, 
and operate on contiguous chunks of memory.  We also use single precision transforms to 
reduce the memory transfer bandwidth and ensure that 
the resampling block can be run on a single device.  The latter point is important since it allows the inout
thread to parallelize the resampling operation across the cluster and provide a roughly x4 speedup.

The first FFT (A) takes the complex valued, 16384 sample SWARM spectrum into the time domain, returning 
32768 real samples.  Since SWARM omits the Nyquist sample which is outside the SWARM guard band cutoff frequency, 
we insert a zero as the 16385th sample.  Since this transform is radix-2, the performance should benefit from 
the Cooley-Tukey algorithm.  However, on our cluster the cuFFT library exhibits linear improvement in Gflops 
with transform size up to N of 16384 (see Figure \ref{fig:C2R_performance}).  This behavior is consistent with 
the CUDA 7.0 performance 
report\footnote{\url{http://on-demand.gputechconf.com/gtc/2015/webinar/gtc-express-cuda7-performance-overview.pdf}} 
and we find that this transform, just outside of the optimal regime, operates at 6.7 Gsps.

@@

Next, we Fourier transform chunks of $N_B = 19968$ consecutive time series samples (B).  The resulting 
spectrum has a frequency resolution of $2496$\,MHz$/19958=125$\,kHz.  The following step is to apply a third
complex-to-real transform (C) on a subset of the spectrum, using only frequencies from $150$\,MHz to $1174$\,MHz.  
The final time series has size $N_C= 2\times(1174-150)/0.125=16384$ per batch element and our desired 
sampling rate of 
$16384 \times 0.125$\,MHz$ = 2048$\,MHz.  This simple operation simultaneously resamples the signal, modulates 
the signal, and trims the guard bands.  Furthermore, it is easily implemented using cuFFT by using the same 
batch size as the second real-to-complex transform, increasing the input pointer index by $150 / 0.125 = 1200$
(equivalent to masking out the first 1200 channels) and then using a input stride of $9980$ for the 
$N=16384$ complex-to-real transform.

@@

\subsection{FFT Accuracy}
One concern with using multiple FFTs to change the sampling rate is the error introduced by bit growth.  The 
floating point error grows as $\mathcal{O}(\sqrt{\log{N}})$ \citep{schatzman96} on average for the Cooley-Tukey
algorithm, but errors can be exacerbated by inaccurate twiddle factors or our choice of FFT algorithm (see
benchFFT\footnote{\url{http://www.fftw.org/accuracy/}}).  We have computed the accuracy of the cuFFT library 
using single precision for 
radix-2 transforms as well as the $N_b = 19968$ case in Figure \ref{fig:fft_accuracy}.  These values 
were computed using benchFFT and we compare against fftw3\footnote{\url{http://www.fftw.org/}}
results also computed on our GPU cluster.  For random input, the cuFFT library follows the expected 
$\mathcal{O}(\sqrt{\log{N}})$ behavior, but with larger scatter and slightly worse performance than fftw3.

@@

APHIDS promises to be an important tool for the EHT.  It crucially enables the correlation in
DIFX of non-standard sampling rates and also provides a platform for any general pre-processing of 
VLBI data.  It features fast data I/O from Mark6 recorders and GPU accelerated processing.  
We have demonstrated the correctness and speed of this design and its immediate applicability to 
the 2015 EHT data.  However,
there remain a number of challenges for future development of the software.

For future VLBI campaigns at the SMA, it would be ideal if APHIDS could run in real-time 
at the backend of the SDBE.  This would reduce the media requirements and also the 
hassle of post-processing the data off-site.  However, the performance will need to improve as the
SWARM data rates increase and if the front-end of the pipeline is optimized to reduce the 
amount of SNR loss from quantization.  Additionally, some analysis on the cooling requirements 
of the cards should be conducted to ensure that the GTX980s used in our GPU cluster can safely run 
at the altitude of the SMA (4100\,meters).  Regardless of where it operates, APHIDS will need 
maintenance to remain compatible with SWARM as it and the SDBE are updated.  The current 
git repository and its wiki is a good solution for keeping track of versioning and performance status.

### 2015-09-04 chapters/conclusion.tex
spell checked

To conclude our work on protoplanary disks, we discussed in Chapter \ref{chap:radial_flows} observational 
indicators of 
radial gas flows within the holes of transition disks.  \citet{casassus15} have studied the gas kinematics for
the curious case of HD142527 using the CO $J$ = $6$-$5$ line and found that their data is consistent with
accretion reaching close to free-fall velocities.  The gas surface density of transition disks have also 
been recently studied using observations taking at sub-millimeter wavelengths 
\citep{bruderer14,zhang14,canovas15,perez15,vandermarel15} and will help determine how these structures are formed
\citep{bruderer13}.  Molecular disk winds, outflows, and binary systems are all recently observed 
with interesting gas kinematics \citep{klaassen13,dutrey14,williams14,salyk14,czekala15}.  Looking to the 
future, well resolved, highly sensitive observations of spectral lines may constrain the strength of MRI-drive
turbulence \citep{simon15}, query the active sites of planet formation \citep{cleeves15,ober15}, or reveal
circumplanetary disks \citep{perez15}.

>>>>

To conclude our work on protoplanetary disks, we discussed in Chapter \ref{chap:radial_flows} observational 
indicators of 
radial gas flows within the holes of transition disks.  \citet{casassus15} have studied the gas kinematics for
the curious case of HD142527 using the CO $J$ = $6$-$5$ line and found that their data is consistent with
accretion reaching close to free-fall velocities.  The gas surface density of transition disks have also 
been recently studied using observations taking at sub-millimeter wavelengths 
\citep{bruderer14,zhang14,canovas15,perez15,vandermarel15} and will help determine how these structures are formed
\citep{bruderer13}.  Molecular disk winds, outflows, and binary systems are all recently observed 
with interesting gas kinematics \citep{klaassen13,dutrey14,williams14,salyk14,czekala15}.  Looking to the 
future, well resolved, highly sensitive observations of spectral lines may constrain the strength of MRI-drive
turbulence \citep{simon15}, query the active sites of planet formation \citep{cleeves15,ober15}, or reveal
circumplanetary disks \citep{perez15}.

### 2015-09-04 chapters/introduction.tex
spell checked

Thomas Young's double split experiment is often a student's first formal introduction to the ideas of 
coherence and interference.  By observing the famous fringe pattern, Young could study the wave-like nature of 
photons and aspects of our physical world that operate on quantum scales.  Today, the techniques of interferometry
are ubiquitious over a wide array of technical fields.  Even within astronomy the scientific applications are 
diverse: interferometers  search for extrasolar planets, find subtle ripples in the early universe and watch 
stars explode.  But in all of these examples the astronomer is searching for small details in our sky that, 
without the resolving power of interferometry, might remain hidden.  Going along with this theme, this thesis 
presents four studies using observations at millimeter wavelenghts that work to reveal fine structures in disks 
around young stars and massive black holes.  We cover a wide variety of techniques, from enabling the production 
of useful data products to the basic interpretation of never before seen features.

@@

Each pair of antennas in an array forms a baseline vector (measured in wavelengths) with length $D_\lambda$ and 
components $(u,v,w)$.  The signal voltages, $V_1(t)$ and $V_2(t)$, measured by the individual 
elements are correlated against each other:
\begin{equation}
r(\tau) = \lim_{T \to \infty} \frac{1}{2T} \int_{-T}^T V_1(t) V_2^\ast(t-\tau) d\tau.
\end{equation}
This operation is often done on site with data streamed directly to a correlator and the resulting 
measurement, or visibility, is stored on disk.  However, for Very Long Baseline Interferometry (VLBI), the 
radio telescopes can be separated by entire continents and real-time correlation is not usually feasible.  For 
these experiments, the signal voltages must be recorded and the visibilities calculated at some later time.  The 
Event Horizon Telescope (EHT) is a leader in the developing field of millimeter VLBI and relies on existing 
observatories.  The difficulty of observing at these short wavelengths increases as the atmosphere becomes 
noiser, time variable, and more opaque.  Consequently, facilities that can operate at these wavelengths are a 
precious resource.

The Submillimeter Array \citep[SMA;][]{ho04} is one such world class array.  Commisioned in 2003, the SMA is 
located on the top of Maunakea on the Big Island of Hawaii at an altitude of 13,386 feet.
It is composed of eight 6 meter diameter dishes which, like many interferometers, can be moved into different 
configurations so that there are elements as close as 15 meters or as far as 500 meters (see Figure 
\ref{fig:sma_alma}).  Its recievers can operate at wavelengths from 430\,microns to 2\,millimeters and 
historically have provided up to 4\,GHz of bandwidth.  Chapter \ref{chap:v4046sgr_structure} features an 
exquisite dataset from the SMA that combines observations taken by multiple configurations.  A major upgrade to 
the array is increasing the bandwidth to 8\,GHz, enabling more sensitive observations and larger spectral 
coverage.  To participate in VLBI experiments, the SMA can operate as a ``phased array'', essentially turning 
the interferometer into a single dish telescope.  Chapter \ref{chap:aphids} addresses a key step of how to 
integrate the SMA into millimeter VLBI observations such as those taken by the EHT.

@@

Over time, this disk of gas and dust will dissipate leaving trace debris: astroids and comets, dust and planets.
So while we may be at the end of the star's formation, this protoplanetary disk is the stage where planets are
born.  Protoplanetary disk composition and structure provides the crucial initial conditions that must 
enable a diverse 
population of planets.  Theories of planet formation are still developing and require the consideration 
of physical processes acting on many physical scales and over a huge range of times.  As such, the
work to test and develop these ideas is quite varied, ranging from theoretical calculations to controlled 
experiments in the lab to the frontiers of space exploration.

@@

The main two ingredients in the disk are gas and dust.  The majority of the disk mass will be in molecular 
hydrogen, but observations focus on easier to detect trace molecules such as $^{12}$CO and its isotopologues.
The dust grains are detected via their bulk thermal emission which, when the emisison is optically thin,  
can be used to measure of dust mass \citep{hildebrand83}.  Interferometric observations taken at 
millimeter wavelengths can resolve structure of nearby protoplanetary disks on scales of a few tens to 
hundreds of astronomical units.  The bulk motion of material in the disk can also be 
studied using these same telescopes by studying how the pattern of line emission
from molecules such as CO changes with frequency \citep{beckwith93}.  The velocity of the gas relative to the 
observer changes as a function of position in the disk. To an observer scanning over frequency, different parts 
of the disk seem to light up when its Doppler shifted emission matches the observer's frequency window.

@@

The Event Horizon Telescope (EHT) observes supermassive black holes with baselines long enough
to resolve event horizon scales of nearby targets.  These VLBI observations provide a 
unique glimpse into one of the most extreme environments in our universe.  Using telescopes in Hawaii, Arizona, 
and California, the collaboration has already detected structure at these scales for Sgr A$^\ast$ in the Galactic 
Center and the black hole at the center of the giant elliptical galaxy M87 
\citep{doeleman08,doeleman09}.  These data provide indirect 
evidence for the existence of an event horizon \citep{broderick15} along with constraints on the degree of 
magnetization in the jet of M87 \citep{kino15} and the orientation of Sgr A$^\ast$ \citep{broderick11}.  EHT 
observations promise to provide tests of general relativity \citep{luminet79,johannsen10,bambi13,broderick14} as 
well as new studies of black hole time variability \citep{doeleman09} and accretion \citep{chan15}.

The bright future of the EHT is partialy driven by its ability to add more telescopes and
incorporate new technologies.  The recent March 2015 observing campaign marked a major milestone in
the EHT expansion with participation from the Large Millimeter Telescope (LMT) in Mexico, Combined Array for 
Research in Millimeter-wave Astronomy (CARMA) in California, the SMA and James Clerk Maxwell Telescope (JCMT) in
Hawaii, SMT in Arizona and the 30m dish at Pico Valeta in Spain.  Figure \ref{fig:eht_baselines} shows the
impressive geographic span of the EHT including sites in Chile and at the South Pole that will soon be 
incorporated into the array.  However, the data taken at the SMA must be 
processed before it
can be correlated with the rest of the array.  In Chapter \ref{chap:aphids} we describe the APHIDS pipeline
which performs a sampling rate conversion using a cluster of Graphcs Processing Units (GPUs).  This software
is intended to be a long-term solution enabling VLBI with the SMA.

>>>>

Thomas Young's double split experiment is often a student's first formal introduction to the ideas of 
coherence and interference.  By observing the famous fringe pattern, Young could study the wave-like nature of 
photons and aspects of our physical world that operate on quantum scales.  Today, the techniques of interferometry
are ubiquitous over a wide array of technical fields.  Even within astronomy the scientific applications are 
diverse: interferometers  search for extrasolar planets, find subtle ripples in the early universe and watch 
stars explode.  But in all of these examples the astronomer is searching for small details in our sky that, 
without the resolving power of interferometry, might remain hidden.  Going along with this theme, this thesis 
presents four studies using observations at millimeter wavelengths that work to reveal fine structures in disks 
around young stars and massive black holes.  We cover a wide variety of techniques, from enabling the production 
of useful data products to the basic interpretation of never before seen features.

@@

Each pair of antennas in an array forms a baseline vector (measured in wavelengths) with length $D_\lambda$ and 
components $(u,v,w)$.  The signal voltages, $V_1(t)$ and $V_2(t)$, measured by the individual 
elements are correlated against each other:
\begin{equation}
r(\tau) = \lim_{T \to \infty} \frac{1}{2T} \int_{-T}^T V_1(t) V_2^\ast(t-\tau) d\tau.
\end{equation}
This operation is often done on site with data streamed directly to a correlator and the resulting 
measurement, or visibility, is stored on disk.  However, for Very Long Baseline Interferometry (VLBI), the 
radio telescopes can be separated by entire continents and real-time correlation is not usually feasible.  For 
these experiments, the signal voltages must be recorded and the visibilities calculated at some later time.  The 
Event Horizon Telescope (EHT) is a leader in the developing field of millimeter VLBI and relies on existing 
observatories.  The difficulty of observing at these short wavelengths increases as the atmosphere becomes 
nosier, time variable, and more opaque.  Consequently, facilities that can operate at these wavelengths are a 
precious resource.

The Submillimeter Array \citep[SMA;][]{ho04} is one such world class array.  Commissioned in 2003, the SMA is 
located on the top of Maunakea on the Big Island of Hawaii at an altitude of 13,386 feet.
It is composed of eight 6 meter diameter dishes which, like many interferometers, can be moved into different 
configurations so that there are elements as close as 15 meters or as far as 500 meters (see Figure 
\ref{fig:sma_alma}).  Its receivers can operate at wavelengths from 430\,microns to 2\,millimeters and 
historically have provided up to 4\,GHz of bandwidth.  Chapter \ref{chap:v4046sgr_structure} features an 
exquisite data set from the SMA that combines observations taken by multiple configurations.  A major upgrade to 
the array is increasing the bandwidth to 8\,GHz, enabling more sensitive observations and larger spectral 
coverage.  To participate in VLBI experiments, the SMA can operate as a ``phased array'', essentially turning 
the interferometer into a single dish telescope.  Chapter \ref{chap:aphids} addresses a key step of how to 
integrate the SMA into millimeter VLBI observations such as those taken by the EHT.

@@

Over time, this disk of gas and dust will dissipate leaving trace debris: asteroids and comets, dust and planets.
So while we may be at the end of the star's formation, this protoplanetary disk is the stage where planets are
born.  Protoplanetary disk composition and structure provides the crucial initial conditions that must 
enable a diverse 
population of planets.  Theories of planet formation are still developing and require the consideration 
of physical processes acting on many physical scales and over a huge range of times.  As such, the
work to test and develop these ideas is quite varied, ranging from theoretical calculations to controlled 
experiments in the lab to the frontiers of space exploration.

@@

The main two ingredients in the disk are gas and dust.  The majority of the disk mass will be in molecular 
hydrogen, but observations focus on easier to detect trace molecules such as $^{12}$CO and its isotopologues.
The dust grains are detected via their bulk thermal emission which, when the emission is optically thin,  
can be used to measure of dust mass \citep{hildebrand83}.  Interferometric observations taken at 
millimeter wavelengths can resolve structure of nearby protoplanetary disks on scales of a few tens to 
hundreds of astronomical units.  The bulk motion of material in the disk can also be 
studied using these same telescopes by studying how the pattern of line emission
from molecules such as CO changes with frequency \citep{beckwith93}.  The velocity of the gas relative to the 
observer changes as a function of position in the disk. To an observer scanning over frequency, different parts 
of the disk seem to light up when its Doppler shifted emission matches the observer's frequency window.

@@

The Event Horizon Telescope (EHT) observes super massive black holes with baselines long enough
to resolve event horizon scales of nearby targets.  These VLBI observations provide a 
unique glimpse into one of the most extreme environments in our universe.  Using telescopes in Hawaii, Arizona, 
and California, the collaboration has already detected structure at these scales for Sgr A$^\ast$ in the Galactic 
Center and the black hole at the center of the giant elliptical galaxy M87 
\citep{doeleman08,doeleman12}.  These data provide indirect 
evidence for the existence of an event horizon \citep{broderick15} along with constraints on the degree of 
magnetization in the jet of M87 \citep{kino15} and the orientation of Sgr A$^\ast$ \citep{broderick11}.  EHT 
observations promise to provide tests of general relativity \citep{luminet79,johannsen10,bambi13,broderick14} as 
well as new studies of black hole time variability \citep{doeleman09} and accretion \citep{chan15}.

The bright future of the EHT is partially driven by its ability to add more telescopes and
incorporate new technologies.  The recent March 2015 observing campaign marked a major milestone in
the EHT expansion with participation from the Large Millimeter Telescope (LMT) in Mexico, Combined Array for 
Research in Millimeter-wave Astronomy (CARMA) in California, the SMA and James Clerk Maxwell Telescope (JCMT) in
Hawaii, SMT in Arizona and the 30m dish at Pico Valeta in Spain.  Figure \ref{fig:eht_baselines} shows the
impressive geographic span of the EHT including sites in Chile and at the South Pole that will soon be 
incorporated into the array.  However, the data taken at the SMA must be 
processed before it
can be correlated with the rest of the array.  In Chapter \ref{chap:aphids} we describe the APHIDS pipeline
which performs a sampling rate conversion using a cluster of Graphics Processing Units (GPUs).  This software
is intended to be a long-term solution enabling VLBI with the SMA.
